---
title: "Bellabeat_case_study"
author: "Edwin King'ori"
date: "`r Sys.Date()`"
output: 
  html_document: 
    number_sections: yes
    toc: yes
    df_print: default
---

<center>Mission Statement</center>

<p style="color:#333333;">

Bellabeat, a high-tech manufacturer of health-focused products for women, believes that analyzing smart device fitness data could help unlock new growth opportunities to become a larger player in the global [smart device] (<https://en.wikipedia.org/wiki/Smart_device>) market. The objective is to focus on one of Bellabeat's products and analyze smart device data to gain insight into how consumers are using their smart devices.The insights discovered will then help guide marketing strategy for the company.

</p>

<center> The Analysis Process.</center>

# ASK.

Analyzing smart device usage data in order to gain insight into how customers use non-bellabeat smart devices.

<strong>1.1 Stakeholders Involved</strong>
<ol style = "list-style: circle">
  <li> Urška Sršen: Bellabeat’s cofounder and Chief Creative Officer.</li>
  <li> Sando Mur: Mathematician and Bellabeat’s cofounder; key member of
  the Bellabeat executive team.</li>
  <li> Bellabeat marketing analytics team: A team of data analysts
  responsible for collecting, analyzing, and reporting data that helps
  guide Bellabeat’s marketing strategy. </li> 
</ol>

<strong>1.2 Insight Questions</strong>
<ol type = "A" >
  <li> What are some trends in smart device usage? </li>  
  <li> How could these trends apply to Bellabeat customers?<?li>
  <li> How could these trends help influence Bellabeat marketing strategy? </li>
</ol>
# Prepare.
  Key Tasks Involved:
  
<strong> 2.1 Identify how it's organized.</strong>
  <ul style=“list-style-type:square” >
  <li> The FitBit Fitness Tracker data set contains personal fitness tracker from thirty fitbit users.</li> 
  <li>Thirty eligible Fitbit users consented to the submission of personal tracker data, including minute-level output for physical activity, heart rate, and sleep monitoring.</li>
    </ul>
      
<strong> 2.3 Determining the credibility of data. </strong>
 <ul style=“list-style-type:square” > 
Is the Data ROCC?  Reliable, Original, Comprehensive, Current, and Cited.
    
  <li> Reliable: Low reliability due to the low number of respondents.</li>
  
  <li> Original:  The data was sourced from a third-party, hence, the source of the data cannot be
    easily verified. </li>
       
  <li> Comprehensive :The data set has 18 csv files with information about daily activity, steps, and
    heart rate that can be used to explore users’ habits. However, It lacks the usre's demographic info. </li>
    
  <li> Time Frame : The data is (8) years old,thus, not relevant. However, we will forego the limitation
    for the purpose of th capstone study. </li> 
       
  <li> Cited: The original [dataset](https://zenodo.org/record/53894) contains the required sources information
  including, publication date, DOI (Document Object Identifier), Authors and file formats.</li> 
  </ul>
       

<strong>2.1 Loading Packages </strong>

```{r}
library(tidyverse)
library(lubridate)
library(skimr)
library(dplyr)
library(janitor)
library(ggplot2)
library(selectr)
```

<strong>2.2 Importing files</strong>

```{r}
activity <- read.csv("dailyActivity_merged.csv")
calories <- read.csv("dailyCalories_merged.csv")
intensities <- read.csv("dailyIntensities_merged.csv")
steps <- read.csv("dailySteps_merged.csv")
sleepday <- read.csv("sleepDay_merged.csv")
Weight <- read.csv("weightLogInfo_merged.csv")

head(activity)
head(calories)
head(intensities)
head(steps)
head(sleepday)
head(Weight)

```

<strong>2.3 Data Cleaning</strong>

2.3.1 Cleaning Column Names

```{r}
activity1 <- janitor::clean_names(activity)
calories1 <- janitor::clean_names(calories)
intensities1 <- janitor::clean_names(intensities)
steps1 <- janitor::clean_names(steps)
sleepday1 <- janitor::clean_names(sleepday)
Weight1 <- janitor::clean_names(Weight)

colnames(activity1)
colnames(calories1)
colnames(intensities1)
colnames(sleepday1)
colnames(steps1)

```
<strong> 2.4 Summarizing data </strong>

Checking sample size to decide whether to use for analysis
<ul style=“list-style-type:square” >
   <li> using the n-distinct() to check the number of unique ID in each data set. </li>
``` {r}
n_distinct(activity$Id)
n_distinct(calories$Id)
n_distinct(intensities$Id)
n_distinct(steps$Id)
n_distinct(sleepday$Id)
n_distinct(Weight$Id)

```
  <li>Using the summary () to get a comprehensive data categorization.</li>
```{r}
activity1 %>%
  select(very_active_minutes,
         fairly_active_minutes,
         lightly_active_minutes) %>%
  summary()
#calories summary
calories1 %>%
  select(calories) %>%
  summary()
#Intensities 
#summarizing the number of active minutes per category
intensities1 %>%
  select(very_active_minutes,
         fairly_active_minutes,
         lightly_active_minutes) %>%
  summary()
#Sleep
sleepday1 %>%
  select(total_sleep_records,
         total_minutes_asleep,
         total_time_in_bed) %>%
  summary()
```

</ul>

2.5 Insights From the Summary
 <ol style = "list-style: circle"> 
 <li> From the distinct data above, 33 participants reported their data for activity, Intensities and Steps. Although only 24 participants shared their data for sleep_day, I'll still use the data set for analysis. </li>
 <li> Only 8 participants partook in the weight log analysis, which is the smallest data set. Thus, will not use it due to bias concerns </li>
 <li> Most participants are lightly active compared to the highly active data.</li>
 <li>The average steps is 7400 per day, which is lower than the recommended 8000 steps</li>
 <li>The average sedentary time is 991, an equivalent of 16 hours which is lower than the medically required 4-8 hours a day according to this [study](https://bmcpublichealth.biomedcentral.com/articles/10.1186/s12889-023-15029-8)</li>
 <li>Most adults spend an approximated 7 hours of sleep. </li>
 </ol>
 
 
 ```{r}
 
```
